---
title: "R01 Sample Size Justification"
author: "J. Lucas McKay"
date: "11/2/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, error = F, message = F)
```

## Data sources

This document is based on data from the file `GaitSignatureQuality.csv` provided by Michael Rosenberg, Ph.D.
This `.csv` file is not tracked by github.

## Sample size justification

We powered the study based on preliminary gait quality data of N = 8 stroke survivors and N = 5 able-bodied comparison subjects.
Across groups, based on 1000 Welch Independent Samples t-tests simulated based on preliminary data with a differential effect size of Cohen's d = 1.3,
a sample size of N = 80 stroke survivors and N = 20 able-bodied individuals will have >99% power to resolve a difference
between stroke and able-bodied.
Within the stroke group, based on contrasts between gait quality measures obtained at enforced speed = 5 and self-selected speed of Cohen's d = 0.8, a paired-sample t-test will have >95% power to resolve an improvement in gait quality accompanying increased gait speed.
Further, based on observed contrasts between enforced speed = 5 and enforced speed = 2 of Cohen's d = 0.5, a paired-sample t-test will have >85% power to resolve differential improvements in gait quality between enforced speeds 5 and 2.


```{r}
library(tidyverse)
d = read_csv("GaitSignatureQuality.csv") |> 
  mutate(Group = case_when(
    grepl("^AB",subjID) ~ "AB",
    T ~ "Stroke"
  )) |> 
  rename(Dist = Mean);

pooled_sd = sd(d$Dist)
# 43

d_mn = d |> group_by(Group,subjID) |> summarize(Dist = mean(Dist))
d_mn_2 = d_mn |> ungroup() |> group_by(Group) |> summarize(Dist_sd = sd(Dist), Dist_mn = mean(Dist))

# calculate cohen's d for stroke vs. able-bodied
cohens_d_1 = diff(d_mn_2$Dist_mn)/pooled_sd
# 1.3

# calculate power at 20/80
sim_ab_st = function(n_ab=20,n_st=80){
  ab = rnorm(n_ab,d_mn_2$Dist_mn[1],d_mn_2$Dist_sd[1])
  st = rnorm(n_st,d_mn_2$Dist_mn[2],d_mn_2$Dist_sd[2])
  m = t.test(ab,st)
  m$p.value
}

p_vals = replicate(1000,sim_ab_st())
power_1 = sum(p_vals<0.05)/length(p_vals)
# [1] 1
# Over 99% power for contrast 1

```
```{r}
# calculate improvement in each speed vs. speed 1
st = d |>
  filter(Group=="Stroke") |>
  group_by(subjID) |>
  mutate(DeltaDist = Dist - Dist[row_number() == 1]) |>
  mutate(Improvement = -DeltaDist) |> 
  select(subjID,SpeedIndex,Improvement) |> 
  mutate(SpeedIndex = factor(SpeedIndex))

st_mn = st |> group_by(SpeedIndex) |> summarize(Improvement_mn = mean(Improvement), Improvement_sd = sd(Improvement))

cohens_d_contrast_2 = st_mn$Improvement_mn[5]/st_mn$Improvement_sd[5]
# [1] 0.8478208

power_2 = power.t.test(n=80,delta=cohens_d_contrast_2)$power
# [1] 0.9996232

cohens_d_contrast_3 = (st_mn$Improvement_mn[5]-st_mn$Improvement_mn[2])/mean(st_mn$Improvement_sd[c(2,5)])
# [1] 0.5034512

power_3 = power.t.test(n=80,delta=cohens_d_contrast_3)$power
# [1] 0.8858459


# ADDITIONAL POWER STATEMENT
# If the main outcomes are resolved further to increase effect size, we can estimate the power to discriminate between all levels of enforced speed.
# with contrasts between levels 2,3,4, and 5, this amounts to 5 post-hoc tests. Performing these with Bonferroni correction will yield alpha = 0.05/5 = 0.01.
# assuming alpha of 0.01,
power.t.test(n=80,power=0.8,sig.level=0.01)
# delta = 0.5460926


```

## Figures

The primary multi-panel data figure is saved as `panels.eps`.
The arguments in the following line can be changed to alter its size and resolution as needed.

```
ggsave("panels.eps",plt,dpi=300,width=6.5,height=3,units="in")
```


## Simple group comparison figure
```{r}
group.colors <- c(AB = "#000040", Stroke = "#800000")
                  
# plot 1
set.seed(585)
f0 = d_mn |> ggplot(aes(Group,Dist,color = Group)) + geom_jitter(width=0.05) +
  geom_point(data = d_mn_2, mapping = aes(Group,Dist_mn, color = Group), shape =5, size = 6) +
  labs(x="",y="Distance from\nAble-bodied Centroid\n(Arbitrary Units)") +
  theme_minimal() +
  theme(legend.position = "none")
ggsave("between.eps",dpi=300,width=2,height=3)

f0
```

## Annotated group comparison figure
```{r}
f1 = d_mn_2 |> ggplot(aes(Group,Dist_mn,color = Group)) + geom_point(shape = 18, size = 6) +
  geom_linerange(aes(ymin = Dist_mn-Dist_sd,ymax = Dist_mn+Dist_sd)) +
  expand_limits(y=c(50,175)) +
  theme_minimal() +
  labs(x="Group",y="Distance from\nAble-bodied Centroid\n(Arbitrary Units)") +
  theme(legend.position = "none") +
  scale_color_manual(values=group.colors) +
  annotate("text", x = 1.5, y = 175, label = "Effect Size\nStroke (N=8)\nvs. AB (N=5)\nCohen's d = 1.3", vjust = "top")
f1
```

## Annotated speed comparison figure
```{r}
# plot 2
set.seed(585)
f2 = st_mn |> 
  mutate(Group = "Stroke") |> 
  ggplot(aes(SpeedIndex,Improvement_mn,color = Group)) + geom_point(shape = 18, size = 6) +
  geom_linerange(aes(ymin = Improvement_mn-Improvement_sd,ymax = Improvement_mn+Improvement_sd)) +
  expand_limits(y = 30) +
  theme_minimal() +
  labs(x="Enforced Speed",y="Improvement towards\nAble-bodied Centroid\n(Arbitrary Units)") +
  theme(legend.position = "none") +
  scale_color_manual(values=group.colors) +
  annotate("text", x = 5, y = 30, label = "Effect Sizes\nSpeed 5 vs. 1: d = 0.8\nSpeed 5 vs. 2: d = 0.5", vjust = "top") +
  annotate("text", x = 1, y = 30, label = "N=8 Stroke", vjust = "top",hjust="left")
f2
```

## Multi-panel figure
```{r}
# gridExtra::grid.arrange(grobs = c(f1,f2), widths = c(1,2), nrow = 1, ncol = 2)
# plt = gridExtra::grid.arrange(f1,f2,ncol = 2)
plt = gridExtra::grid.arrange(f1,f2,widths=c(1.5,2))
ggsave("panels.eps",plt,dpi=300,width=6.5,height=3,units="in")

```

